---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
Please note, that to implement the plotting in this report, the `ggplot2` library has been used. For all code snippets below, this library is already available.
```{r echo = FALSE, results = "hide"}
library(ggplot2)
Sys.setlocale(category = "LC_ALL", locale = "en_US.UTF-8")
```

## Loading and preprocessing the data
To load the data into the R environment, I am going to take the contents from the Github branch as a starting point, which includes the activity monitoring data inside a zip-compressed file. 

The following steps have to be performed from there:

1.  extract the core data source file, by using the unzip function in R
2.  ingest the data using read.csv to read the file, taking into account the type of each data column, NA values, and format of dates
3.  clean up things not needed anymore. I by purpose remove unneeded objects, and also remove intermittent files 

The respective code looks like this:
```{r tidy=TRUE}
AMD.zipname <- "activity.zip"
AMD.filename <- "activity.csv"
# unpack the original source file
if (!file.exists(AMD.filename)) {
  unzip(zipfile =  AMD.zipname, files = AMD.filename)
}

# ingest data source into data.table with appropriate column types
AMD.data.raw <- read.csv(AMD.filename,
                     header = TRUE,
                     sep = ",",
                     colClasses = c(steps="numeric", date="Date", interval="numeric"), na.strings = "NA")
```
This will internalize the data into a variable called AMD.data, with the following structure:
```{r echo=FALSE}
str(AMD.data.raw)
```

## What is mean total number of steps taken per day?
To calculate the mean total number of steps per day, we need to aggregate the steps of all intervals for each individual day. The result shall be shown in histogram.

The code looks like this:
```{r tidy=TRUE}
## aggregating the data
AMD.aggregated.raw <- aggregate(steps ~ date, data = AMD.data.raw, sum)

## creating the histogram plot
qplot(AMD.aggregated.raw$steps, 
      geom = "histogram", 
      binwidth = 1000,
      main = "Overview on the frequency of total steps per day on the raw data (incl. NA)", 
      xlab = "Number of steps",
      ylab = "Number of days with specific steps")
```

The mean and median of the frequency distribution can be easily explored like this:
```{r tidy = TRUE}
summary(AMD.aggregated.raw$steps)
```

## What is the average daily activity pattern?
By calculating the average number of steps per interval over all days, we could see whether there are specific points during the day, were there are more activities than during other times. This does not take into account, that the day during the week (i.e. weekdays and weekends) most probably might create a difference, too.

Anyway, we average over the intervals, which looks like this:
```{r tidy = TRUE}
## aggregated per interval over all days
AMD.msi.raw <- aggregate(steps ~ interval, data = AMD.data.raw, mean)

# creating the line plot
qplot(interval, steps, 
      data = AMD.msi.raw, 
      geom = c("line"),
      main = "Average Number of Steps per Interval (over all Days)", 
      xlab = "Interval",
      ylab = "Mean Number of Steps")
```

## Imputing missing values

### Identify incomplete observations
The `NA` value could generally be observed in each variable of an observation, in which case it is not complete. Therefore, I can use the `complete.cases` function to select the proper cases.

On the other hand, I looked through the variables separately, and realized, that only the `steps` variable will have `NA` values.

```{r tidy = TRUE}
cases.quality <- complete.cases(AMD.data.raw)
cases.nok <- sum(!cases.quality)

nrow(AMD.data.raw[is.na(AMD.data.raw$steps),])
nrow(AMD.data.raw[is.na(AMD.data.raw$date),])
nrow(AMD.data.raw[is.na(AMD.data.raw$interval),])
```

We see that we have `r cases.nok` of incomplete cases, of a total number of `r nrow(AMD.data.raw)`.

My strategy to replace the `NA` values will be to assign the mean for the 5-minute interval to it.
Since I already have identified the incomplete cases (see above), I need to iterate over them and assign the values from `AMD.msi`.

```{r tidy = TRUE}
# Create a copy of the data set, in order to keep the original
AMD.data.cleaned <- AMD.data.raw

# Cycle through those rows in the set, which have an incomplete observation
# Assign the mean (calculated above) to the 'steps' variable of that row 
for (row in which(!cases.quality)) {
  i <- AMD.data.cleaned[row, c("interval")]
  AMD.data.cleaned[row, c("steps")] <- AMD.msi.raw[AMD.msi.raw$interval == i, c("steps")]
}
```

### Histogram of the Cleaned Dataset
Creating the new histogram is straight forward and identical to the previous one, only this time using `AMD.data.cleaned`.
```{r tidy=TRUE}
## aggregating the data
AMD.aggregated.cleaned <- aggregate(steps ~ date, data = AMD.data.cleaned, sum)

## creating the histogram plot
qplot(AMD.aggregated.cleaned$steps, 
      geom = "histogram", 
      binwidth = 1000,
      main = "Overview on the frequency of total steps per day on the cleaned data", 
      xlab = "Number of steps",
      ylab = "Number of days with specific steps")
```

### Comparison of the Mean and Median
Comparing the two analysis, is straight forward:
```{r tidy = TRUE}
summary(AMD.aggregated.raw$steps)
summary(AMD.aggregated.cleaned$steps)
```
We can see that the impact is the mean staying unchanged, while the median has shifted slightly upward, as for 1st and 3rd quantiles.

## Differences in Activity Patterns Weekdays vs. Weekends?
The patterns were analyzed on the mean number of steps per interval of all days. We will now create a factor variable to identify those dates which are either weekdays or weekend.
```{r tidy = TRUE}
# attach a now variable to the data set, with a factor "weekday/weekend". Use the date variable to derive it.
AMD.data.raw <- cbind(AMD.data.raw, 
                      weekday = ifelse(!weekdays(AMD.data.raw$date, abbreviate = TRUE) %in% c("Sat", "Sun"), "weekday", "weekend"))

# Create the mean steps per interval dataset, including the weekday factor
AMD.msi.raw <- aggregate(steps ~ interval + weekday, data = AMD.data.raw, mean)

# Create the plot, clustered by the 'weekday' variable
qplot(interval, steps, 
      data = AMD.msi.raw,
      geom = c("line"),
      facets = weekday ~ .,      
      main = "Average Number of Steps per Interval (Clustered by Weekday vs. Weekend)", 
      xlab = "Interval",
      ylab = "Mean Number of Steps")
```
The intervals on weekdays have less number of steps on an average in some ranges of the day, while weekend days show more activity distributed throughout the day. As we can see, the weekend has a higher mean activity, than the week.
```{r tidy = TRUE}
mean(AMD.msi.raw[AMD.msi.raw$weekday == "weekday",]$steps)
mean(AMD.msi.raw[AMD.msi.raw$weekday == "weekend",]$steps)
```

## Conclusion
A conclusion we might take is that the analyzed person has a job with lots of sitting, or similar, and that he uses his weekends to get active. So this could be the pattern of a office worker or student. It would be interesting to compare more activity measurements from those two groups side by side, in order to be more precise in guessing the occupation.

Furthermore, it is interesting to see that early in the day there is a peak in the data, regardless of whether we are analyzing weekdays or weekends. This might indicate some special activity during that period, e.g. going out for sports.
